---
title: "Classification"
subtitle: "Logistic regression, thresholding, ROC/AUC, and k-nearest neighbors"
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

From regression to classification
===

In regression, the response is quantitative.

In classification, the response is categorical.

Examples:

- email is spam or not spam,
- loan defaults or does not default,
- tumor is benign or malignant.

For classification, we often want:

- a **predicted class**,
- and sometimes a **predicted probability**.

Example: the `Default` data
===

We will use the `Default` data from `ISLR2`.

- response: `default` (`Yes` or `No`)
- predictors: `balance` and `income`

We will compare:

- **logistic regression**
- **k-nearest neighbors (kNN) classification**

Train/test split
===

```{r data-setup, eval=FALSE}
library(ISLR2)

data(Default)

dat <- Default[, c("default", "balance", "income")]

y <- dat$default
x <- as.matrix(dat[, c("balance", "income")])

set.seed(1)
n <- nrow(dat)
train_id <- sample(seq_len(n), size = round(n / 2))
test_id <- setdiff(seq_len(n), train_id)
```

Logistic regression
===

Logistic regression models the conditional probability of class membership:

\[
p(x) = P(Y = 1 \mid X = x).
\]

Instead of modeling \(p(x)\) directly, it models the **log-odds**:

\[
\log\left(\frac{p(x)}{1 - p(x)}\right)
=
\beta_0 + \beta_1 x_1 + \cdots + \beta_p x_p.
\]

This guarantees fitted probabilities between 0 and 1.

Fit logistic regression in `R`
===

```{r logistic-fit, eval=FALSE}
log_fit <- glm(default ~ balance + income,
               data = dat,
               subset = train_id,
               family = binomial)

log_prob <- predict(log_fit,
                    newdata = dat[test_id, ],
                    type = "response")

head(log_prob)
range(log_prob)
```

Logistic regression gives probabilities
===

A fitted probability is more informative than a class label alone.

Examples:

- 0.02 suggests a very low estimated chance of default,
- 0.45 suggests much more uncertainty,
- 0.90 suggests a very high estimated chance of default.

To get a class label, we still need to choose a **threshold**.

From probabilities to class labels
===

A common default threshold is 0.5:

```{r logistic-class, eval=FALSE}
log_class <- ifelse(log_prob > 0.5, "Yes", "No")
log_class <- factor(log_class, levels = levels(y))

table(predicted = log_class,
      truth = y[test_id])

mean(log_class != y[test_id])
```

The threshold does not have to be 0.5.

Confusion matrix
===

For a binary classifier, the confusion matrix records:

- **TP**: predicted Yes, truth Yes
- **FP**: predicted Yes, truth No
- **TN**: predicted No, truth No
- **FN**: predicted No, truth Yes

These counts help us move beyond a single overall error rate.


True positive rate and false positive rate
===

Two important summaries are:

\[
\text{TPR} = \frac{TP}{TP + FN}
\]

and

\[
\text{FPR} = \frac{FP}{FP + TN}.
\]

Interpretation:

- **TPR** = among the true Yes observations, how many did we correctly flag?
- **FPR** = among the true No observations, how many did we incorrectly flag?


Threshold choice changes the classifier
===

A smaller threshold usually:

- predicts **Yes** more often,
- increases **TPR**,
- but also increases **FPR**.

A larger threshold usually:

- predicts **Yes** less often,
- decreases **FPR**,
- but may also decrease **TPR**.

This is a tradeoff.

Calibration: are the probabilities meaningful?
===

A model is **well calibrated** if observations with predicted probability near \(p\) are positive about \(p\) fraction of the time.

Examples:

- among cases with predicted probability around 0.10, about 10% should be Yes,
- among cases with predicted probability around 0.70, about 70% should be Yes.

Calibration is about the **probabilities**, not the thresholded class labels.

ROC curve
===

The **ROC curve** tracks classifier performance across many thresholds.

For each threshold, compute:

- x-axis: **FPR**
- y-axis: **TPR**

So the ROC curve summarizes the TPR/FPR tradeoff over all thresholds.

Compute ROC points manually
===

```{r roc-points, eval=FALSE}
threshold_grid <- seq(0, 1, by = 0.01)
roc_mat <- matrix(NA, nrow = length(threshold_grid), ncol = 2)
colnames(roc_mat) <- c("FPR", "TPR")

for (i in seq_along(threshold_grid)) {
  th <- threshold_grid[i]
  pred_i <- factor(ifelse(log_prob > th, "Yes", "No"),
                   levels = levels(y))

  TP_i <- sum(pred_i == "Yes" & y[test_id] == "Yes")
  FP_i <- sum(pred_i == "Yes" & y[test_id] == "No")
  TN_i <- sum(pred_i == "No"  & y[test_id] == "No")
  FN_i <- sum(pred_i == "No"  & y[test_id] == "Yes")

  roc_mat[i, ] <- c(FP_i / (FP_i + TN_i),
                    TP_i / (TP_i + FN_i))
}

plot(roc_mat[, "FPR"], roc_mat[, "TPR"], type = "l",
     xlab = "False positive rate",
     ylab = "True positive rate")
abline(0, 1, lty = 2)
```

AUC
===

The **area under the ROC curve (AUC)** summarizes the ROC curve with one number.

- AUC near 0.5: weak discrimination
- AUC near 1: strong discrimination

Interpretation:

- AUC is the probability that a randomly chosen Yes case receives a higher score than a randomly chosen No case.

Approximate AUC with the trapezoid rule
===

```{r auc-manual, eval=FALSE}
ord_roc <- order(roc_mat[, "FPR"], roc_mat[, "TPR"])
roc_ord <- roc_mat[ord_roc, , drop = FALSE]

auc <- sum(diff(roc_ord[, "FPR"]) *
             (head(roc_ord[, "TPR"], -1) + tail(roc_ord[, "TPR"], -1)) / 2)

auc
```

How should we evaluate a classifier?
===

Different summaries answer different questions:

- **test error / accuracy**: how often are labels correct at one threshold?
- **TPR / FPR**: what kinds of mistakes are we making at that threshold?
- **calibration**: do predicted probabilities behave like probabilities?
- **ROC / AUC**: how well does the model rank observations across thresholds?

k-nearest neighbors classification
===

kNN classification works very differently.

To classify a new point:

1. find the \(k\) nearest training observations,
2. look at their class labels,
3. use the majority vote.

The tuning parameter is \(k\).

- small \(k\): more flexible, lower bias, higher variance
- large \(k\): smoother, higher bias, lower variance

Why preprocessing matters for kNN
===

kNN uses distances, so the scale of each predictor matters.

If one predictor is measured on a much larger scale, it can dominate the distance calculation.

So for kNN we usually standardize the predictors.

**Important:** the means and standard deviations must be computed from the training data only.

```{r scale-knn, eval=FALSE}
mu <- colMeans(x[train_id, , drop = FALSE])
s  <- apply(x[train_id, , drop = FALSE], 2, sd)

x_train <- scale(x[train_id, , drop = FALSE], center = mu, scale = s)
x_test  <- scale(x[test_id,  , drop = FALSE], center = mu, scale = s)
```

Manual CV loop to choose \(k\)
===

We will tune \(k\) by 5-fold CV on the training data.

```{r knn-cv, eval=FALSE}
library(class)

set.seed(11)

K <- 5
fold_id <- sample(rep(1:K, length.out = length(train_id)))
k_grid <- c(1, 3, 5, 7, 9, 11, 15, 21, 31, 51)

cv_err_knn <- matrix(NA, nrow = K, ncol = length(k_grid))

for (fold in 1:K) {

  x_tr_raw <- x[train_id, , drop = FALSE][fold_id != fold, , drop = FALSE]
  y_tr <- y[train_id][fold_id != fold]

  x_val_raw <- x[train_id, , drop = FALSE][fold_id == fold, , drop = FALSE]
  y_val <- y[train_id][fold_id == fold]

  mu_fold <- colMeans(x_tr_raw)
  s_fold  <- apply(x_tr_raw, 2, sd)

  x_tr <- scale(x_tr_raw, center = mu_fold, scale = s_fold)
  x_val <- scale(x_val_raw, center = mu_fold, scale = s_fold)

  for (j in seq_along(k_grid)) {
    pred_fold <- knn(train = x_tr, test = x_val, cl = y_tr, k = k_grid[j])
    cv_err_knn[fold, j] <- mean(pred_fold != y_val)
  }
}

knn_cv_mean <- colMeans(cv_err_knn)
k_hat <- k_grid[which.min(knn_cv_mean)]
```

Fit the final kNN classifier
===

```{r knn-final, eval=FALSE}
knn_pred <- knn(train = x_train,
                test = x_test,
                cl = y[train_id],
                k = k_hat)

table(predicted = knn_pred,
      truth = y[test_id])

knn_test_error <- mean(knn_pred != y[test_id])

k_hat
knn_test_error
```

Compare logistic regression and kNN
===

```{r compare-classifiers, eval=FALSE}
log_test_error <- mean(log_class != y[test_id])

c(Logistic = log_test_error,
  kNN = knn_test_error)
```

Summary
===

- Classification predicts categorical outcomes.
- Logistic regression returns probabilities, then we choose a threshold for class labels.
- A confusion matrix helps us compute TPR, FPR, and other threshold-specific summaries.
- Calibration asks whether predicted probabilities behave like probabilities.
- ROC curves and AUC summarize discrimination across thresholds.
- kNN uses a tuning parameter \(k\) and depends strongly on preprocessing.
- Standardization for kNN should be done using training data only.
- Cross-validation can be used to tune \(k\) before final test-set evaluation.
