---
title: "Regularized Regression"
subtitle: "Ridge regression, the lasso, and cross-validation"
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Why do we need regularization?
===

The least squares linear regression model is arguably the **most widely used** statistical model and has a place in nearly every application domain of statistics. It works well across a vast range of settings. That said, it can be unstable for prediction when:

- we have many predictors,
- predictors are strongly correlated,
- or the fitted model is too sensitive to small changes in the data.

In those settings, linear regression may fit the training data very well but generalize less well to new observations.

Regularization addresses this by adding a **penalty** that shrinks the coefficient estimates. The underlying statistical concept is to inject *bias* into the model, in the hopes of reducing *variance*. 

Ridge regression
===

Recall that least squares minimizes

\[
\mathrm{RSS} = \sum_{i=1}^n \left(y_i - \beta_0 - \sum_{j=1}^p \beta_j x_{ij}\right)^2.
\]

Ridge regression minimizes

\[
\mathrm{RSS} + \lambda \sum_{j=1}^p \beta_j^2,
\]

where \(\lambda \ge 0\) is a tuning parameter.

As \(\lambda\) increases:

- the coefficients are pulled toward 0,
- the model becomes more stable,
- variance tends to decrease,
- bias tends to increase.

The lasso
===

The lasso uses a different penalty:

\[
\mathrm{RSS} + \lambda \sum_{j=1}^p |\beta_j|.
\]

Like ridge regression, the lasso shrinks coefficients toward 0.

But unlike ridge regression, the lasso can set some coefficients **exactly equal to 0**.

That means the lasso can perform:

- regularization for prediction,
- and model selection for a simpler model.

How will we choose \(\lambda\)?
===

Here we will **fit the model with `glmnet()`**, but we will implement the **cross-validation loop ourselves**.

For a grid of candidate \(\lambda\) values:

1. split the training data into folds,
2. hold out one fold at a time,
3. fit the model on the remaining folds,
4. compute MSE on the held-out fold,
5. average the MSE across folds,
6. choose the \(\lambda\) with the smallest mean CV error.

Example setup: the `Credit` data
===

We will use the `Credit` data from `ISLR2`.

- response: `Balance`
- predictors: the remaining variables

We use `model.matrix()` because the data contain both quantitative and qualitative predictors - this is a quick way of converting your categorical predictors into dummy variables. 

```{r data-setup, eval=FALSE}
library(glmnet)
library(ISLR2)

data(Credit)

X <- model.matrix(Balance ~ ., data = Credit)[, -1]
y <- Credit$Balance

set.seed(1)
n <- nrow(X)
train_id <- sample(1:n, size = round(n / 2))

x_train <- X[train_id, , drop = FALSE]
y_train <- y[train_id]

x_test <- X[-train_id, , drop = FALSE]
y_test <- y[-train_id]
```

Create the folds and the tuning grid
===

We will use 5-fold CV on the **training set only**.

```{r folds-grid, eval=FALSE}
set.seed(11)

K <- 5
fold_id <- sample(rep(1:K, length.out = length(train_id)))

lambda_grid <- 10^seq(4, -2, length.out = 100)
```

Manual CV loop for ridge
===

We still fit each ridge model with `glmnet()`, but the **cross-validation logic** is ours (base R only).

```{r ridge-cv-loop, eval=FALSE}
cv_mse_ridge <- matrix(NA, nrow = K, ncol = length(lambda_grid))

for (k in 1:K) {

  x_tr <- x_train[fold_id != k, , drop = FALSE]
  y_tr <- y_train[fold_id != k]

  x_val <- x_train[fold_id == k, , drop = FALSE]
  y_val <- y_train[fold_id == k]

  ridge_k <- glmnet(x_tr, y_tr,
                    alpha = 0,
                    lambda = lambda_grid,
                    standardize = TRUE)

  pred_k <- predict(ridge_k, newx = x_val, s = lambda_grid)

  cv_mse_ridge[k, ] <-
    colMeans((matrix(y_val,
                     nrow = length(y_val),
                     ncol = length(lambda_grid)) - pred_k)^2)
}

ridge_cv_mean <- colMeans(cv_mse_ridge)
ridge_lambda_hat <- lambda_grid[which.min(ridge_cv_mean)]
```

Select \(\lambda\) and refit ridge on the full training set
===

```{r ridge-refit, eval=FALSE}
ridge_final <- glmnet(x_train, y_train,
                      alpha = 0,
                      lambda = ridge_lambda_hat,
                      standardize = TRUE)

ridge_pred <- as.numeric(predict(ridge_final, newx = x_test,
                                 s = ridge_lambda_hat))

ridge_test_mse <- mean((y_test - ridge_pred)^2)

ridge_lambda_hat
ridge_test_mse
```

Manual CV loop for lasso
===

The lasso uses exactly the same CV structure, but with `alpha = 1`.

```{r lasso-cv-loop, eval=FALSE}
cv_mse_lasso <- matrix(NA, nrow = K, ncol = length(lambda_grid))

for (k in 1:K) {

  x_tr <- x_train[fold_id != k, , drop = FALSE]
  y_tr <- y_train[fold_id != k]

  x_val <- x_train[fold_id == k, , drop = FALSE]
  y_val <- y_train[fold_id == k]

  lasso_k <- glmnet(x_tr, y_tr,
                    alpha = 1,
                    lambda = lambda_grid,
                    standardize = TRUE)

  pred_k <- predict(lasso_k, newx = x_val, s = lambda_grid)

  cv_mse_lasso[k, ] <-
    colMeans((matrix(y_val,
                     nrow = length(y_val),
                     ncol = length(lambda_grid)) - pred_k)^2)
}

lasso_cv_mean <- colMeans(cv_mse_lasso)
lasso_lambda_hat <- lambda_grid[which.min(lasso_cv_mean)]
```

Refit lasso and inspect the selected coefficients
===

```{r lasso-refit, eval=FALSE}
lasso_final <- glmnet(x_train, y_train,
                      alpha = 1,
                      lambda = lasso_lambda_hat,
                      standardize = TRUE)

lasso_pred <- as.numeric(predict(lasso_final, newx = x_test,
                                 s = lasso_lambda_hat))

lasso_test_mse <- mean((y_test - lasso_pred)^2)

coef(lasso_final)
lasso_test_mse
```

Compare test MSE
===

We can compare linear regression, ridge, and lasso on the held-out test set.

```{r compare-test-mse, eval=FALSE}
reg_fit <- lm(Balance ~ ., data = Credit, subset = train_id)

reg_pred <- predict(reg_fit, newdata = Credit[-train_id, ])
reg_mse <- mean((y_test - reg_pred)^2)

c(LS = reg_mse,
  Ridge = ridge_test_mse,
  Lasso = lasso_test_mse)
```

Summary
===

- Ridge and lasso are regularized versions of linear regression. 
- Both introduce a tuning parameter \(\lambda\).
- We can let `glmnet()` fit the model while we code the CV loop ourselves.
- The key computational idea is to separate **model fitting** from **tuning logic**.
- Final model comparison should be based on test-set performance.

