---
title: 'Statistical Learning Setup and Cross-Validation'
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR2)
```

From fitting to prediction
===

So far, we have focused on **fitting** a regression model. 

We have mostly utilized the `lm()` function to fit a linear regression model. We use this model to do inference or understand the relationship between a response $Y$ and a set of covariates (collectively denoted as $X$).

But in many problems, especially for modern applications, the real goal is **prediction**:

- fit a model on data we already have
- use that model to predict outcomes for **new observations**
- evaluate the quality of those predictions. 

In a statistical computing class, this raises a practical question: how do we assess a model's ability to predict new observations? 

A simple regression example
===

We will use the `Auto` data and predict `mpg` from `horsepower`.

So our starting model is

\[
\texttt{mpg} \sim \texttt{horsepower}.
\]

```{r}
library(ISLR2)

dat <- na.omit(Auto[, c("mpg", "horsepower")])
n <- nrow(dat)

plot(dat$horsepower, dat$mpg,
     pch = 19,
     col = "gray40",
     xlab = "horsepower",
     ylab = "mpg",
     main = "Auto data: mpg vs horsepower")

fit_full <- lm(mpg ~ horsepower, data = dat)
abline(fit_full, lwd = 2, col = 2)
```

Prediction error and MSE
===

Suppose we fit a model and get predicted values \(\widehat y_i\).

A common way to measure prediction error is the **mean squared error**:

$$
\text{MSE} = \frac{1}{m}\sum_{i=1}^m (y_i - \widehat y_i)^2.
$$

### Interpretation

- large errors are penalized more heavily because of the square
- smaller MSE means predictions are, on average, closer to the observed outcomes
- for regression, MSE is one of the most common prediction metrics

Training MSE
===

The **training data** are the observations used to fit the model.

If we compute MSE on those same observations, we get the **training MSE**.

$$
\text{Training MSE} =
\frac{1}{n_{\text{train}}}
\sum_{i \in \text{train}} (y_i - \widehat y_i)^2.
$$

### Important point

Training MSE is usually **too optimistic** as a measure of future prediction performance,
because the model has already seen those data, therefore by construction it is designed to fit that data well. 

Compute the training MSE
===

```{r}
fit_train_only <- lm(mpg ~ horsepower, data = dat)
pred_train <- predict(fit_train_only, newdata = dat)

train_mse <- mean((dat$mpg - pred_train)^2)
train_mse
```

Why is this not enough?
===

A low training MSE does **not** guarantee good performance on new data.

Why not?

- the coefficients were chosen using these same observations
- the model is being evaluated on data it was fit to
- future observations will not be identical to the training sample

For prediction, the quantity we really care about is closer to:

### Real target

How well does the fitted model predict **new, unseen data**?

Training set and test set
===

A basic solution is the **train/test split**.

We divide the data into two parts:

- **training set**: used to fit the model
- **test set**: held out and used only for evaluation

Then the **test MSE** is

$$
\text{Test MSE} =
\frac{1}{n_{\text{test}}}
\sum_{i \in \text{test}} (y_i - \widehat y_i)^2,
$$

where the predictions come from the model fit on the training set.

One random train/test split
===

```{r}
set.seed(1)

test_id <- sample(1:n, size = round(n/2), replace = FALSE)
train_dat <- dat[-test_id, ]
test_dat  <- dat[test_id, ]

fit_split <- lm(mpg ~ horsepower, data = train_dat)

pred_train <- predict(fit_split, newdata = train_dat)
pred_test  <- predict(fit_split, newdata = test_dat)

train_mse <- mean((train_dat$mpg - pred_train)^2)
test_mse  <- mean((test_dat$mpg - pred_test)^2)

c(train_mse = train_mse, test_mse = test_mse)
```

How should we interpret this?
===

Typically:

- the **training MSE** is smaller
- the **test MSE** is larger

That is not by coincidence. 

Tell me why. 

### Takeaway

For prediction, **test error** is usually much more informative than training error.

A weakness of one split
===

The validation-set approach is simple, but it has a weakness:

### The answer depends on the split.

If a different group of observations lands in the test set, we can get a different test MSE.

So one split can be:

- noisy
- unstable
- sensitive to random partitioning

Repeated random splits
===

We can see this instability by repeating the train/test split many times.

```{r}
set.seed(2)

B <- 100
split_mse <- numeric(B)

for (b in 1:B) {
  test_id <- sample(1:n, size = round(0.25 * n), replace = FALSE)
  train_dat <- dat[-test_id, ]
  test_dat  <- dat[test_id, ]
  
  fit_b <- lm(mpg ~ horsepower, data = train_dat)
  pred_b <- predict(fit_b, newdata = test_dat)
  
  split_mse[b] <- mean((test_dat$mpg - pred_b)^2)
}

head(split_mse)
mean(split_mse)
sd(split_mse)
```

Visualizing the split-to-split variability
===

```{r}
hist(split_mse,
     breaks = 15,
     main = "Test MSE across repeated random splits",
     xlab = "test MSE",
     col = "gray80",
     border = "white")
```

### Interpretation

The single train/test split gives **one estimate** of prediction error.
Repeated splits show that this estimate can vary quite a bit.

Cross-validation
===

Cross-validation is a more systematic way to estimate prediction error.

The idea is to:

- split the data into several parts called **folds**
- hold out one fold at a time
- fit the model on the remaining folds
- evaluate prediction error on the held-out fold
- average the fold-specific errors

This uses the data more efficiently than a single train/test split.

K-fold cross-validation
===

In **K-fold CV**:

1. divide the observations into \(K\) folds
2. for fold \(k\), fit the model on all observations not in fold \(k\)
3. predict the observations in fold \(k\)
4. compute the validation error for fold \(k\)
5. average the \(K\) fold errors

$$
\text{CV}_K = \frac{1}{K}\sum_{k=1}^K \text{Err}^{(k)}.
$$

Common choices are \(K=5\) or \(K=10\).

Create fold labels in base R
===

```{r}
set.seed(3)

K <- 5
fold_id <- sample(rep(1:K, length.out = n))

table(fold_id)
head(fold_id)
```

Each observation gets a fold label.

If `fold_id[i] = 3`, then observation `i` is used as validation data when we hold out fold 3.

K-fold CV by hand: one loop
===

```{r}
cv_mse <- numeric(K)

for (k in 1:K) {
  train_dat <- dat[fold_id != k, ]
  valid_dat <- dat[fold_id == k, ]
  
  fit_k <- lm(mpg ~ horsepower, data = train_dat)
  pred_k <- predict(fit_k, newdata = valid_dat)
  
  cv_mse[k] <- mean((valid_dat$mpg - pred_k)^2)
}

cv_mse
mean(cv_mse)
```

What the loop is doing
===

Each pass through the loop does three things:

- defines the current training and validation sets
- refits the regression model on the training portion only
- evaluates prediction error on the held-out fold


### Statistical computing view

Cross-validation is just an **algorithm**:

- partition the indices
- refit repeatedly
- store the errors
- average at the end

From evaluating one model to comparing several
===

So far, we have used cross-validation to estimate prediction error for **one** model.

But this is also how we compare **candidate models**.

For example, we might compare:

- a linear model
- a quadratic model
- a cubic model

### Main rule

Use the **same folds** for all models, then compare their average validation errors.

Candidate models for comparison
===

Using the same response and predictor, consider:

$$
\begin{aligned}
M_1 &: \texttt{mpg} \sim \texttt{horsepower} \\
M_2 &: \texttt{mpg} \sim \texttt{poly(horsepower, 2)} \\
M_3 &: \texttt{mpg} \sim \texttt{poly(horsepower, 3)}
\end{aligned}
$$

These models have increasing flexibility.

The question is not:

### Which one has the smallest training MSE?

The prediction-focused question is:

### Which one has the smallest estimated **out-of-sample** error?

Comparing models with the same CV folds
===

```{r}
cv_mse_1 <- numeric(K)
cv_mse_2 <- numeric(K)
cv_mse_3 <- numeric(K)

for (k in 1:K) {
  train_dat <- dat[fold_id != k, ]
  valid_dat <- dat[fold_id == k, ]
  
  fit_1 <- lm(mpg ~ horsepower, data = train_dat)
  fit_2 <- lm(mpg ~ poly(horsepower, 2), data = train_dat)
  fit_3 <- lm(mpg ~ poly(horsepower, 3), data = train_dat)
  
  pred_1 <- predict(fit_1, newdata = valid_dat)
  pred_2 <- predict(fit_2, newdata = valid_dat)
  pred_3 <- predict(fit_3, newdata = valid_dat)
  
  cv_mse_1[k] <- mean((valid_dat$mpg - pred_1)^2)
  cv_mse_2[k] <- mean((valid_dat$mpg - pred_2)^2)
  cv_mse_3[k] <- mean((valid_dat$mpg - pred_3)^2)
}

c(mean(cv_mse_1), mean(cv_mse_2), mean(cv_mse_3))
```

Summarize the comparison
===

```{r}
cv_summary <- data.frame(
  model = c("linear", "quadratic", "cubic"),
  cv_mse = c(mean(cv_mse_1), mean(cv_mse_2), mean(cv_mse_3))
)

cv_summary[order(cv_summary$cv_mse), ]
```

### Interpretation

The model with the smallest average CV error is the one that predicts best **among these candidates**, according to this CV procedure.

This is still a **prediction** argument, not a hypothesis test.

Later, the same idea will be used for tuning more flexible procedures like ridge and lasso.

LOOCV as a special case
===

If \(K=n\), each fold contains exactly one observation.

This is called **leave-one-out cross-validation (LOOCV)**.

### Idea

For each observation:

- leave that one point out
- fit the model on the other \(n-1\) observations
- predict the left-out point
- square the error

Then average those \(n\) squared errors.

### In practice

LOOCV is conceptually simple, but 5-fold or 10-fold CV is often used because it is computationally cheaper and usually more stable.
