---
title: 'Control Flow: apply() family'
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

The apply family
===

R offers a family of **apply functions**, which allow you to apply a function across different chunks of data. Offers an alternative to explicit iteration using `for()` loop; can be simpler and faster, **though not always** and can lack readability. Summary of functions:

- `apply()`: apply a function to rows or columns of a matrix or data frame
- `lapply()`: apply a function to elements of a list or vector
- `sapply()`: same as the above, but simplify the output (if possible)
- `tapply()`: apply a function to levels of a factor vector

`apply()`: rows or columns of a matrix or data frame
===

The `apply()` function takes inputs of the following form:

- `apply(x, MARGIN=1, FUN=my.fun)`, to apply `my.fun()` across rows of a matrix or data frame `x`
- `apply(x, MARGIN=2, FUN=my.fun)`, to apply `my.fun()` across columns of a matrix or data frame `x`

```{r}
out <- numeric(nrow(X))

for (i in 1:nrow(X)) {
  out[i] <- mean(X[i, ])
}

out <- apply(X, 1, mean)
```

`apply()` is still a loop. You just don’t see the counter.

```{r}
class(state.x77) # Built-in matrix of states data, 50 states x 8 variables
head(state.x77) 

apply(state.x77, MARGIN=2, FUN=min) # Minimum entry in each column
apply(state.x77, MARGIN=2, FUN=max) # Maximum entry in each column
apply(state.x77, MARGIN=2, FUN=which.max) # Index of the max in each column
```

Applying a custom function 
===

For a custom function, we can just define it before hand, and the use `apply()` as usual

```{r}
# Our custom function: trimmed mean
trimmed.mean = function(v) {  
  q1 = quantile(v, prob=0.1)
  q2 = quantile(v, prob=0.9)
  return(mean(v[q1 <= v & v <= q2]))
}

apply(state.x77, MARGIN=2, FUN=trimmed.mean) 
```

We'll learn more about functions later (don't worry too much at this point about the details of the function definition)

Applying a custom function "on-the-fly"
===

Instead of defining a custom function before hand, we can just define it "on-the-fly". Sometimes this is more convenient

```{r}
# Compute trimmed means, defining this on-the-fly
apply(state.x77, MARGIN=2, FUN=function(v) { 
  q1 = quantile(v, prob=0.1)
  q2 = quantile(v, prob=0.9)
  return(mean(v[q1 <= v & v <= q2]))
})
```

Applying a function that takes extra arguments
===

Can tell `apply()` to pass **extra arguments** to the function in question. E.g., can use: `apply(x, MARGIN=1, FUN=my.fun, extra.arg.1, extra.arg.2)`, for two extra arguments `extra.arg.1`, `extra.arg.2` to be passed to `my.fun()`

```{r}
# Our custom function: trimmed mean, with user-specified percentiles
trimmed.mean = function(v, p1, p2) {
  q1 = quantile(v, prob=p1)
  q2 = quantile(v, prob=p2)
  return(mean(v[q1 <= v & v <= q2]))
}

apply(state.x77, MARGIN=2, FUN=trimmed.mean, p1=0.01, p2=0.99)
```

What's the return argument?
===

What kind of data type will `apply()` give us? Depends on what function we pass. Summary, say, with `FUN=my.fun()`:

- If `my.fun()` returns a single value, then `apply()` will return a vector
- If `my.fun()` returns k values, then `apply()` will return a matrix with k rows (note: this is true *regardless* of whether `MARGIN=1` or `MARGIN=2`)
- If `my.fun()` returns different length outputs for different inputs, then `apply()` will return a list
- If `my.fun()` returns a list, then `apply()` will return a list

We'll grapple with this on the lab. This is one main advantage of `purrr` package: there is a much more transparent return object type

Optimized functions for special tasks
===

**Don't overuse** the apply paradigm! There's lots of special functions that **optimized** are will be both simpler and faster than using `apply()`. E.g.,

- `rowSums()`, `colSums()`: for computing row, column sums of a matrix
- `rowMeans()`, `colMeans()`: for computing row, column means of a matrix
- `max.col()`: for finding the maximum position in each row of a matrix

Combining these functions with logical indexing and vectorized operations will enable you to do quite a lot. E.g., how to count the number of positives in each row of a matrix?

```{r}
x = matrix(rnorm(9), 3, 3)
# Don't do this (much slower for big matrices)
apply(x, MARGIN=1, function(v) { return(sum(v > 0)) })
# Do this insted (much faster, simpler)
rowSums(x > 0)
```

<span style="color: blue">Part III</span>
===

<span style="color: blue">*`lapply()`, `sapply()`, `tapply()`*</span>


`lapply()`: elements of a list or vector
===

The `lapply()` function takes inputs as in: `lapply(x, FUN=my.fun)`, to apply `my.fun()` across elements of a list or vector `x`. The output is always a list

```{r}
my.list
lapply(my.list, FUN=mean) # Get a warning: mean() can't be applied to chars
lapply(my.list, FUN=summary)
```

`sapply()`: elements of a list or vector
===

The `sapply()` function works just like `lapply()`, but tries to **simplify** the return value whenever possible. E.g., most common is the conversion from a list to a vector

```{r}
sapply(my.list, FUN=mean) # Simplifies the result, now a vector
sapply(my.list, FUN=summary) # Can't simplify, so still a list
```

`tapply()`: levels of a factor vector
===

The function `tapply()` takes inputs as in: `tapply(x, INDEX=my.index, FUN=my.fun)`, to apply `my.fun()` to subsets of entries in `x` that share a common level in `my.index`

```{r}
# Compute the mean and sd of the Frost variable, within each region
tapply(state.x77[,"Frost"], INDEX=state.region, FUN=mean)
tapply(state.x77[,"Frost"], INDEX=state.region, FUN=sd)
```

`split()`: split by levels of a factor
===

The function `split()` split up the rows of a data frame by levels of a factor, as in: `split(x, f=my.index)` to split a data frame `x` according to levels of `my.index`

```{r}
# Split up the state.x77 matrix according to region
state.by.reg = split(data.frame(state.x77), f=state.region)
class(state.by.reg) # The result is a list
names(state.by.reg) # This has 4 elements for the 4 regions
class(state.by.reg[[1]]) # Each element is a data frame
```

---

```{r}
# For each region, display the first 3 rows of the data frame
lapply(state.by.reg, FUN=head, 3) 
```

---

```{r}
# For each region, average each of the 8 numeric variables
lapply(state.by.reg, FUN=function(df) { 
  return(apply(df, MARGIN=2, mean)) 
})
```

Summary
===

- Data frames are a representation of the "classic" data table in R: rows are observations/cases, columns are variables/features
- Each column can be a different data type (but must be the same length)
- `subset()`: function for extracting rows of a data frame meeting a condition 
- `split()`: function for splitting up rows of a data frame, according to a factor variable
- `apply()`: function for applying a given routine to rows or columns of a matrix or data frame
- `lapply()`: similar, but used for applying a routine to elements of a vector or list
- `sapply()`: similar, but will try to simplify the return type, in comparison to `lapply()`
- `tapply()`: function for applying a given routine to groups of elements in a vector or list, according to a factor variable