---
title: 'Tidyverse I: Tibbles, Pipes, and dplyr Verbs'
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Tidyverse
===

The **tidyverse** is a collection of R packages designed to make data analysis workflows readable and consistent.

Today we focus on three pieces:

- **Tibbles and tidy data**
- **The pipe**
- **Core `dplyr` verbs** for data wrangling

```{r, eval=FALSE}
library(tidyverse)
```

If you do not have the tidyverse installed:

```{r, eval=FALSE}
install.packages("tidyverse")
```

Why tidyverse?
===

For statistical computing, tidyverse is useful because it gives us a readable grammar for moving from raw data to analysis-ready data.

A common workflow is:

1. Import or create data
2. Clean and organize the data
3. Create summaries
4. Make plots
5. Fit models
6. Evaluate results

The goal is not to replace base R. The goal is to learn another syntax that is especially convenient for data analysis pipelines.

A Reproducibility Caution
===

The tidyverse is widely used and very helpful for readable data analysis code, but it is built from external R packages that change over time.

Tidyverse code is often readable and efficient, but because tidyverse packages evolve over time, long-term reproducibility requires more attention to package versions than base R code.

For example, older code may use `gather()` and `spread()`, while newer code usually uses `pivot_longer()` and `pivot_wider()`.

This does **not** mean we should avoid the tidyverse. It means that for reproducible work, we should be cautious. 

Base R functions change more slowly and are always available with R itself, while tidyverse functions require additional packages.

Example data
===

We will use a small dataset so that we can see exactly what each function does.

```{r}
library(tidyverse)

students <- tibble(
  id = 1:8,
  section = c("A", "A", "A", "B", "B", "B", "B", "A"),
  hours = c(5, 8, 3, 6, 9, 2, 4, 7),
  score = c(82, 91, 74, 88, 95, 70, 79, 86),
  passed = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE)
)

students
```

Tibbles
===

A **tibble** is a modern version of a data frame.

```{r}
students
```

Tibbles behave very similarly to data frames, but print more neatly and are often easier to inspect.

Useful functions:

```{r}
glimpse(students)
```

Tibble vs. data frame
===

Base R data frame:

```{r}
students_df <- data.frame(
  id = 1:8,
  section = c("A", "A", "A", "B", "B", "B", "B", "A"),
  hours = c(5, 8, 3, 6, 9, 2, 4, 7),
  score = c(82, 91, 74, 88, 95, 70, 79, 86),
  passed = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE)
)

students_df
```

Tidyverse version:

```{r}
students_tbl <- tibble(
  id = 1:8,
  section = c("A", "A", "A", "B", "B", "B", "B", "A"),
  hours = c(5, 8, 3, 6, 9, 2, 4, 7),
  score = c(82, 91, 74, 88, 95, 70, 79, 86),
  passed = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE)
)

students_tbl
```

Converting between data frames and tibbles
===

You can convert a data frame to a tibble with `as_tibble()`.

```{r}
mtcars_tbl <- as_tibble(mtcars)
mtcars_tbl
```

Base R version: `data.frame()` creates a standard data frame.

```{r}
mtcars_df <- data.frame(mtcars)
head(mtcars_df)
```

Both are acceptable data structures for most statistical modeling functions in R.

Tidy data
===

A dataset is usually called **tidy** when:

- each row is an observation
- each column is a variable
- each cell contains one value

For example, this is tidy:

```{r}
students
```

Each row is one student. Each column is one variable about that student.

The pipe
===

The pipe sends the object on the left into the function on the right.

```{r}
students %>%
  filter(section == "A")
```

Read this as:

> Take `students`, then keep rows where `section == "A"`.

The pipe is mainly about "readability".

Without the pipe
===

This code:

```{r}
students %>%
  filter(section == "A")
```

is equivalent to:

```{r}
students_df[students_df$section=="A",]
```

The pipe becomes more useful when we want to do several steps in sequence.

With several steps
===

```{r}
students %>%
  filter(section == "A") %>%
  select(id, hours, score) %>%
  arrange(desc(score))
```

Read this as:

1. Start with `students`
2. Keep section A
3. Keep only selected columns
4. Sort by score from highest to lowest

The pipe helps the code follow the order of the analysis.

One equivalent version in base R: 
```{r}
# Keep only students in section A
students_A <- students_df[students_df$section == "A", ]

# Keep selected columns
students_A <- students_A[, c("id", "hours", "score")]

# Sort by score from highest to lowest
students_A <- students_A[order(-students_A$score), ]

students_A
```

Core dplyr verbs
===

Some important `dplyr` verbs are:

- `filter()` keeps rows
- `select()` keeps columns
- `mutate()` creates or modifies columns
- `arrange()` sorts rows
- `summarize()` collapses data into summaries
- `group_by()` applies operations within groups

Today we start with the first four.

`filter()`: keep rows
===

Use `filter()` to keep rows that satisfy a condition.

```{r}
students %>%
  filter(score >= 85)
```

Base R version:

```{r}
students[students$score >= 85, ]
```

Both versions return the students with score at least 85.

Multiple filter conditions
===

Use `&` for “and”.

```{r}
students %>%
  filter(section == "B" & score >= 80)
```

Base R version:

```{r}
students[students$section == "B" & students$score >= 80, ]
```

Use `|` for “or”.

```{r}
students %>%
  filter(score >= 90 | hours >= 8)
```

Base R version:

```{r}
students[students$score >= 90 | students$hours >= 8, ]
```

`select()`: keep columns
===

Use `select()` to keep selected variables.

```{r}
students %>%
  select(id, section, score)
```

Base R version:

```{r}
students[, c("id", "section", "score")]
```

This does not change the rows. It changes which columns we keep.

`mutate()`: create or modify columns
===

Use `mutate()` to create a new variable.

```{r}
students %>%
  mutate(score_percent = score / 100)
```

Base R version:

```{r}
students_base <- students
students_base$score_percent <- students_base$score / 100
students_base
```

The original rows are preserved. We have added a new column.

Creating indicator variables
===

A common use of `mutate()` is creating an indicator variable.

```{r}
students %>%
  mutate(high_score = score >= 90)
```

Base R version:

```{r}
students_base <- students
students_base$high_score <- students_base$score >= 90
students_base
```


Using `if_else()`
===

We can also create a labeled variable.

```{r}
students %>%
  mutate(performance = if_else(score >= 85, "high", "not high"))
```

Base R version:

```{r}
students_base <- students
students_base$performance <- ifelse(students_base$score >= 85, "high", "not high")
students_base
```

Note: `if_else()` is the tidyverse version. `ifelse()` is the base R version.

`arrange()`: sort rows
===

Use `arrange()` to sort rows by a variable.

```{r}
students %>%
  arrange(score)
```

Base R version:

```{r}
students[order(students$score), ]
```

By default, sorting is from smallest to largest.

Descending order
===

Use `desc()` inside `arrange()` for descending order.

```{r}
students %>%
  arrange(desc(score))
```

Base R version:

```{r}
students[order(students$score, decreasing = TRUE), ]
```

Combining verbs
===

The tidyverse style is especially helpful when we combine several data-wrangling steps.

```{r}
students %>%
  filter(passed == TRUE) %>%
  mutate(score_percent = score / 100) %>%
  select(id, section, hours, score_percent) %>%
  arrange(desc(score_percent))
```

The code reads from top to bottom in the same order as the workflow.

Base R version of the same workflow
===

```{r}
students_base <- students[students$passed == TRUE, ]
students_base$score_percent <- students_base$score / 100
students_base <- students_base[, c("id", "section", "hours", "score_percent")]
students_base <- students_base[order(students_base$score_percent, decreasing = TRUE), ]
students_base
```

This does the same thing. The difference is mostly syntax and what you consider readability.

Combining verbs to create grouped summaries
===

We can also combine verbs to answer a summary question.

Question: among students who passed, what is the average score in each section?

```{r}
students %>%
  filter(passed == TRUE) %>%
  group_by(section) 
```

Base R version:

```{r}
students_passed <- students[students$passed == TRUE, ]

students_by_section <- split(students_passed, students_passed$section)
students_by_section[["A"]]
```
Combining verbs to create grouped summaries
===

We can also combine verbs to answer a summary question.

Question: among students who passed, what is the average score in each section?

```{r}
students %>%
  filter(passed == TRUE) %>%
  group_by(section) %>%
  summarize(
    mean_score = mean(score)
  ) %>%
  arrange(desc(mean_score))
```

Base R version:

```{r}
passed_students <- students[students$passed == TRUE, ]
passed_summary <- aggregate(score ~ section, data = passed_students, FUN = mean)

names(passed_summary)[which(names(passed_summary)=="score")] <- "mean_score"
passed_summary
```

Keep average for every student
===

Compute a group level summary but without collapsing the data: 

```{r}
students <- students %>%
  group_by(section) %>%
  mutate(section_mean = mean(score)) %>%
  ungroup() #removes the grouping structure after the grouped calculation is finished, good habit.

students
```

Base R version: 
```{r}
students$section_mean <- ave(
  students$score,
  students$section,
  FUN = mean
)

students
```


Main takeaways
===

- A **tibble** is a modern data frame.
- **Tidy data** means rows are observations and columns are variables.
- The **pipe** lets us write workflows from left to right and top to bottom.
- `filter()` keeps rows.
- `select()` keeps columns.
- `mutate()` creates or modifies columns.
- `arrange()` sorts rows.
- Most tidyverse operations have direct base R analogues.

