---
title: 'Tidyverse II: Creating Analysis-Ready Variables'
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Today: analysis-ready variables
===

Last time, we introduced:

- tibbles
- the pipe
- `filter()`
- `select()`
- `mutate()`
- `arrange()`
- `group_by()` and `summarize()`

Today we will use tidyverse tools to create variables and summaries that are useful for analysis.

The goal is practical:

> How do we turn raw variables into variables that answer statistical questions?

Load packages
===

```{r}
library(tidyverse)
```

We will use a small dataset so that the code is easy to inspect.

```{r}
students <- tibble(
  id = 1:10,
  section = c("A", "A", "A", "B", "B", "B", "B", "A", "A", "B"),
  hours = c(5, 8, 3, 6, 9, 2, 4, 7, 10, 1),
  score = c(82, 91, 74, 88, 95, 70, 79, 86, 97, 65),
  passed = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, FALSE)
)

students
```

Why create new variables?
===

Raw variables are not always in the form we need for analysis.

For example, we may want to create:

- an indicator for whether a student passed
- a label for high versus low performance
- a category such as low, medium, or high
- standardized versions of several variables
- group-level summaries

This is one of the most common uses of `mutate()`.

Review: `mutate()`
===

`mutate()` creates a new column or modifies an existing column.

```{r}
students %>%
  mutate(score_percent = score / 100)
```

Base R version:

```{r}
students_base <- students
students_base$score_percent <- students_base$score / 100
students_base
```

Creating TRUE/FALSE variables
===

Sometimes we want a logical variable.

```{r}
students %>%
  mutate(high_score = score >= 90)
```

Base R version:

```{r}
students_base <- students
students_base$high_score <- students_base$score >= 90
students_base
```

This creates a variable that is either `TRUE` or `FALSE`.

Creating labels with `if_else()`
===

Use `if_else()` when there are two possible outcomes.

```{r}
students %>%
  mutate(
    performance = if_else(score >= 85, "high", "not high")
  )
```

Read this as:

> If `score >= 85`, label the student as `"high"`. Otherwise, label the student as `"not high"`.

Base R version: `ifelse()`
===

The base R version is `ifelse()`.

```{r}
students_base <- students
students_base$performance <- ifelse(
  students_base$score >= 85,
  "high",
  "not high"
)

students_base
```

The idea is the same:

```{r, eval=FALSE}
ifelse(condition, value_if_true, value_if_false)
```

Using more than one condition
===

We can combine conditions using `&` for "and".

```{r}
students %>%
  mutate(
    strong_and_studied = if_else(score >= 85 & hours >= 6,
                                 "yes", "no")
  )
```

Base R version:

```{r}
students_base <- students
students_base$strong_and_studied <- ifelse(
  students_base$score >= 85 & students_base$hours >= 6,
  "yes",
  "no"
)

students_base
```

Using `case_when()`
===

Use `case_when()` when there are more than two categories.

```{r}
students %>%
  mutate(
    score_group = case_when(
      score < 75 ~ "low",
      score >= 75 & score < 90 ~ "medium",
      score >= 90 ~ "high"
    )
  )
```

This is easier to read than many nested `if_else()` statements.

How to read `case_when()`
===

The structure is:

```{r, eval=FALSE}
case_when(
  condition_1 ~ value_1,
  condition_2 ~ value_2,
  condition_3 ~ value_3
)
```

Each line means:

> If this condition is true, use this value.

The order matters. R checks the conditions from top to bottom.

Base R version: nested `ifelse()`
===

One base R version uses nested `ifelse()`.

```{r}
students_base <- students
students_base$score_group <- ifelse(
  students_base$score < 75,
  "low",
  ifelse(students_base$score < 90, "medium", "high")
)

students_base
```

This works, but it can become hard to read when there are many categories.

A default category
===

It is often useful to include a default category using `TRUE ~`.

```{r}
students %>%
  mutate(
    attendance_group = case_when(
      hours < 4 ~ "low hours",
      hours >= 4 & hours < 8 ~ "moderate hours",
      hours >= 8 ~ "high hours",
      TRUE ~ "other"
    )
  )
```

`TRUE ~ "other"` catches anything that did not match the earlier conditions.

Checking your new variables
===

After creating a new variable, check it.

```{r}
students2 <- students %>%
  mutate(
    score_group = case_when(
      score < 75 ~ "low",
      score >= 75 & score < 90 ~ "medium",
      score >= 90 ~ "high"
    )
  )

table(students2$score_group)
```

Base R checking function:

```{r}
table(students_base$score_group)
```

`count()`: a tidyverse shortcut
===

The tidyverse also has `count()`.

```{r}
students2 %>%
  count(score_group)
```

This is equivalent to grouping and summarizing:

```{r}
students2 %>%
  group_by(score_group) %>%
  summarize(n = n())
```

Base R version:

```{r}
table(students2$score_group)
```

Motivation for `across()`
===

Sometimes we want to apply the same operation to several columns.

For example, suppose we want the mean of:

- `hours`
- `score`

We could write:

```{r}
students %>%
  summarize(
    mean_hours = mean(hours),
    mean_score = mean(score)
  )
```

But this becomes repetitive when there are many variables.

Using `across()` inside `summarize()`
===

`across()` lets us apply the same function to multiple columns.

```{r}
students %>%
  summarize(
    across(c(hours, score), mean)
  )
```

Read this as:

> Summarize across the columns `hours` and `score` by taking the mean.

Base R version:

```{r}
sapply(students[, c("hours", "score")], mean)
```

Using `across()` with groups
===

`across()` is especially useful with `group_by()`.

```{r}
students %>%
  group_by(section) %>%
  summarize(
    across(c(hours, score), mean)
  )
```

Base R version:

```{r}
aggregate(
  students[, c("hours", "score")],
  by = list(section = students$section),
  FUN = mean
)
```

Naming the output
===

We can control the names of the new summary columns.

```{r}
students %>%
  group_by(section) %>%
  summarize(
    across(c(hours, score), mean, .names = "mean_{.col}")
  )
```

Here, `{.col}` is replaced by the original column name.

So `score` becomes `mean_score`.

Using `across()` with missing values
===

Many real datasets have missing values.

```{r}
students_missing <- students
students_missing$score[c(3, 6)] <- NA
students_missing
```

We can use an anonymous function inside `across()`.

```{r}
students_missing %>%
  summarize(
    across(c(hours, score), ~ mean(.x, na.rm = TRUE))
  )
```

Base R version:

```{r}
sapply(students_missing[, c("hours", "score")], mean, na.rm = TRUE)
```

Using `across()` inside `mutate()`
===

We can also use `across()` to create new versions of several columns.

For example, create standardized versions of `hours` and `score`.

```{r}
students %>%
  mutate(
    across(
      c(hours, score),
      ~ (.x - mean(.x)) / sd(.x),
      .names = "{.col}_z"
    )
  )
```

This keeps the original variables and adds new variables.

Base R version
===

```{r}
students_base <- students
students_base$hours_z <- (students_base$hours - mean(students_base$hours)) / sd(students_base$hours)
students_base$score_z <- (students_base$score - mean(students_base$score)) / sd(students_base$score)
students_base
```

The base R version is completely fine.

The tidyverse version is useful when we want to repeat the same operation across several variables.

Putting it together
===

Question:

> Among students who passed, what are the average hours studied and average score by score group?

```{r}
students %>%
  mutate(
    score_group = case_when(
      score < 75 ~ "low",
      score >= 75 & score < 90 ~ "medium",
      score >= 90 ~ "high"
    )
  ) %>%
  filter(passed == TRUE) %>%
  group_by(score_group) %>%
  summarize(
    across(c(hours, score), mean, .names = "mean_{.col}"),
    n = n()
  )
```

Main takeaways
===

- `mutate()` creates or modifies variables.
- `if_else()` is useful for two categories.
- `case_when()` is useful for three or more categories.
- Always check newly created variables.
- `across()` applies the same operation to multiple columns.
- The tidyverse is useful when the workflow becomes easier to read from top to bottom.
