---
title: 'Jackknife and Bias Estimation'
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR2)
```

From bootstrap to jackknife
===

So far, we have used the **bootstrap** to approximate a sampling distribution by resampling **with replacement**.

The **jackknife** takes a different approach.

Instead of drawing many bootstrap samples, we:

- leave out one observation at a time
- recompute the statistic each time
- use those leave-one-out values to study variability and bias

### Main idea

The jackknife is a deterministic resampling method based on **systematic deletion**.

What is the jackknife?
===

Suppose our statistic of interest is $\widehat{\theta}$, computed from \(n\) observations.

For each \(i = 1,\dots,n\), let

\[
\widehat{\theta}_{(-i)} = \text{the statistic computed after removing observation } i.
\]

This gives \(n\) leave-one-out estimates.

Let

\[
\bar{\theta}_{(\cdot)} = \frac{1}{n}\sum_{i=1}^n \widehat{\theta}_{(-i)}.
\]

These values are the basis for jackknife estimates of **bias** and **standard error**.

Jackknife bias estimate
===

A standard jackknife estimate of bias is

\[
\widehat{\mathrm{bias}}_{\text{jack}} = (n-1)\big(\bar\theta_{(\cdot)} - \widehat\theta\big).
\]

This comes from a second-order term in a Taylor-type expansion,. 

A jackknife bias-corrected estimator is

\[
\widehat\theta_{\text{jack}} = \widehat\theta - \widehat{\mathrm{bias}}_{\text{jack}}.
\]

### Interpretation

- if the estimated bias is positive, \(\widehat\theta\) tends to be too large
- if the estimated bias is negative, \(\widehat\theta\) tends to be too small
- bias correction subtracts off that estimated bias

Jackknife standard error
===

A standard jackknife estimate of standard error is

$$
\widehat{\mathrm{SE}}_{\text{jack}} = 
\left(\tfrac{n-1}{n}\sum_{i=1}^n \left(\widehat\theta_{(-i)} - \bar\theta_{(\cdot)}\right)^2 \right)^{1/2}
$$

### Takeaway

The jackknife can be used for:

- **bias estimation**
- **bias correction**
- **standard error estimation**

It tends to work best for **smooth statistics**.


### Takeaway

Both methods estimate bias, but the bootstrap does so by resampling **with replacement**, whereas the jackknife uses **leave-one-out** samples.

Jackknife example: standard error for correlation
===

We can use the jackknife to estimate a standard error.

Using the `Auto` data, let

\[
\widehat\theta = \text{corr}(\texttt{horsepower}, \texttt{mpg}).
\]

Then we leave out one row at a time and recompute the sample correlation.

Jackknife SE in R
===

```{r}
data(Auto)

dat <- Auto[, c("horsepower", "mpg")]
n <- nrow(dat)

theta_hat <- cor(dat$horsepower, dat$mpg)

theta_loo <- numeric(n)

for (i in 1:n) {
  dat_i <- dat[-i, ]
  theta_loo[i] <- cor(dat_i$horsepower, dat_i$mpg)
}

theta_bar <- mean(theta_loo)

se_jack <- sqrt((n - 1) / n * sum((theta_loo - theta_bar)^2))

theta_hat
se_jack
```

Jackknife vs bootstrap
===

### Jackknife

- systematic leave-one-out procedure
- deterministic once the data are fixed
- often useful for **bias estimation**
- often works well for **smooth statistics**
- a local sensitivity / influence calculation
- It asks: How much does the estimator change when one data point is removed?

### Bootstrap

- resample with replacement 
- simulation-based (randomness)
- more flexible in complicated settings
- often preferred for confidence intervals and general uncertainty quantification
- It asks: What would happen if the data-generating process were approximated by the empirical distribution?

What is my inferential goal?
===

### If I want:

- a quick SE estimate
- a bias estimate
- an influence-style diagnostic

--> then jackknife is attractive.

### If I want:

- a finite-sample empirical sampling distribution
- confidence intervals
- a more flexible procedure that generalizes across many settings

--> then bootstrap is usually better.