---
title: 'Rejection Sampling and Importance Sampling'
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

Where we are in the course
===

So far, we have mostly relied on:

- Built-in samplers like `rnorm()`, `rpois()`, `runif()`, ...
- **Inverse transform sampling** (Inverse CDF method): works great **when you can compute the inverse CDF** $F^{-1}(u)$.

But in practice, you often face one (or more) of these problems:

1. You *can evaluate a density* $f(x)$, but you **cannot invert** the CDF.
2. The event you care about is **rare**, so naive simulation wastes most samples.
3. You need your sample to **represent important subgroups** (or reduce Monte Carlo variance).

Today: three practical tools for these situations.

Outline
===

1. Rejection sampling (to *generate* data when no inverse CDF is available)
2. Importance sampling (to *estimate* expectations/probabilities efficiently, especially rare events)
3. Stratified sampling (to improve precision / ensure representation)

A quick decision guide (preview)
===

**Goal: Generate *independent* draws from a distribution**

- If you know $F^{-1}$: **Inverse transform**
- If you *can bound* $f(x)$ by a scaled proposal: **Rejection sampling**


**Goal: Estimate $E[h(X)]$ or $P(X \in A)$**

- If $A$ is rare under your simulator: **Importance sampling**
- If there are natural subgroups (regions, age groups, machines, bins): **Stratified sampling**

You can combine methods (e.g., stratify + importance sampling within each stratum).

Rejection sampling: when and why
===

**When to use it**

You want to simulate from a target density $f(x)$ but:

- $F^{-1}$ is not available (no closed-form quantile function)
- you can still evaluate $f(x)$ up to a constant, i.e. $f(x) \propto \tilde f(x)$

**Key idea**

Sample from an easier proposal distribution $g(x)$, then *accept* a draw with some probability so that accepted values follow the target.

Rejection sampling: the algorithm 
===

Assume:

- You can sample from a proposal density $g(x)$
- You can compute an unnormalized target $\tilde f(x)$
- You can find a constant $M$ such that for all $x$ in the support:
  \[
  \tilde f(x) \le M g(x).
  \]

Algorithm:

1. Draw $Y \sim g$
2. Draw $U \sim \text{Uniform}(0,1)$
3. Accept $Y$ if
   \[
   U \le \frac{\tilde f(Y)}{M g(Y)}.
   \]
4. Repeat until you have $n$ accepted draws

**Practical note:** The acceptance rate is about $1/M$ (higher is better).

Rejection sampling example: a “weird” density with no inverse CDF
===

Target density on $[-2,2]$:
\[
f(x) \propto \exp(-x^4), \quad -2 \le x \le 2.
\]

- This is a valid density after normalization (but we don't need the normalizing constant).
- The CDF has no simple closed form, so inverse transform is not convenient.
- Easy proposal: Uniform$(-2,2)$.

Because $\tilde f(x) = \exp(-x^4) \le 1$ and $g(x)=1/4$, we can use $M=4$.

Rejection sampler in base R
===

```{r, echo=TRUE}
set.seed(4279)

f_tilde <- function(x) exp(-x^4)                # unnormalized target
g_samp  <- function(n) runif(n, min=-2, max=2)  # proposal sampler
g_dens  <- function(x) dunif(x, min=-2, max=2)  # proposal density

rej_samp <- function(n, M = 4) {
  out <- numeric(0)
  n_try <- 0

  while (length(out) < n) {
    y <- g_samp(1)
    u <- runif(1)
    n_try <- n_try + 1

    # acceptance probability = f_tilde(y) / (M * g(y))
    # Here it simplifies to exp(-y^4).
    if (u <= f_tilde(y) / (M * g_dens(y))) out <- c(out, y)
  }

  list(draws = out, acceptance_rate = n / n_try)
}

sim <- rej_samp(5000)
sim$acceptance_rate
```

Visual check: histogram vs (scaled) density
===

```{r, echo=TRUE, fig.height=4, fig.width=8}
x <- seq(-2, 2, length.out = 400)

hist(sim$draws, breaks = 40, freq = FALSE, col = "lightgray",
     main = "Rejection Sampling: draws from f(x) ∝ exp(-x^4)",
     xlab = "x")

# Overlay a *scaled* version of the unnormalized density (for shape only)
curve(f_tilde(x) / integrate(f_tilde, -2, 2)$value, add = TRUE, lwd = 2, col = "blue")
```

Rejection sampling: practical tips and pitfalls
===

**Choosing the proposal $g$**

- A *bad* proposal (too narrow, wrong location) forces a large $M$ → very low acceptance
- A *good* proposal looks like the target (same support, similar shape)

**Red flags in practice**

- Acceptance rate is tiny (e.g., < 1%)
- You keep rejecting values in certain regions → proposal mismatch

**Takeaway:** Rejection sampling is great for generating data, but can become inefficient in higher dimensions or with poor proposals.


Importance sampling: when and why
===

**When to use it**

You want to estimate an expectation or probability under a target distribution, e.g.

- $E_f[h(X)]$
- $P_f(X \in A)$

and naive simulation is inefficient, especially if:

- $A$ is a *rare event*
- $h(X)$ is dominated by tail behavior

**Key idea**

Simulate from a different distribution that visits “important” regions more often, and then reweight.

Importance sampling: the rare-event problem
===

Suppose we care about a **very rare event**, like:

- a system failure that happens about **1 in 10,000** runs
- an extreme loss in finance
- a “once in a long time” tail outcome

If we simulate *normally* (from the usual model), almost every simulated run is a **non-failure**.

That means:

- we spend lots of compute/time generating “boring” samples
- our estimate of failure probability is noisy because we see few (or zero) failures

Importance sampling: the basic idea
===

Instead of sampling under “typical conditions,” we:

1. **Intentionally over-sample high-risk conditions** where failures are more likely  
2. **Down-weight** those high-risk samples so we don’t exaggerate the overall risk

So we get:

- **more information** about the rare event per simulation run
- a **much faster** (often far less variable) estimate of the true risk

A helpful mental picture
===

- Standard simulation: “look for a needle in a haystack”
- Importance sampling: “sample from a smaller pile where needles are more common”

But after we collect needles from the smaller pile, we correct for the fact that we **looked in a biased place** by applying **weights**.

Importance sampling: what is a weight?
===

A **weight** is a correction factor that says:

> “How much should this sample count when estimating the original probability?”

- Samples drawn from **high-risk** conditions usually get **small weights**
- Samples drawn from **typical** conditions get **larger weights**

You can think of weights as undoing the bias introduced by “oversampling risk.”

**Our Goal**: estimate an average or probability under the original model \(p(x)\):
\[
\mu = \mathbb{E}_{p}[f(X)].
\]

We simulate from a different model \(q(x)\) and correct with weights:
\[
\mu
= \int f(x)\,p(x)\,dx
= \int f(x)\,\frac{p(x)}{q(x)}\,q(x)\,dx
= \mathbb{E}_{q}\left[f(X)\,\underbrace{\frac{p(X)}{q(X)}}_{\text{weight}}\right].
\]

That’s all: **sample from \(q\)**, then **reweight** to estimate under \(p\).

Practical guidance: choosing the “high-risk sampler”
===

Your “high-risk” sampling distribution should:

- put more probability where the rare event happens (so you actually see it)
- still cover all the important regions of the original model

Rule of thumb for tail events:

- If the proposal has **lighter tails** than the original, weights can become **huge**
- Huge weights → one or two samples dominate → estimates become unstable

Common pitfall (what can go wrong)
===

**Bad choice:** your high-risk sampler avoids some region where the original model can go.

Result:

- weights can “explode”
- variance can get **worse** than standard Monte Carlo

In practice, you should always check:

- Do a few weights dominate?
- Does the estimate jump around a lot when you rerun?


Importance sampling example: estimating a rare event probability
===

Target: \(Z \sim N(0,1)\). We want
\[
p = P(Z > 4).
\]

This is rare (about 3e-5), so with naive sampling you might see **zero** events unless \(n\) is huge.

Proposal: \(Z \sim N(\mu, 1)\) with \(\mu=4\), so we generate more tail values near 4.

Naive Monte Carlo vs importance sampling
===

```{r, echo=TRUE}
set.seed(4279)
n <- 50000
threshold <- 4

# True value (for checking only)
p_true <- 1 - pnorm(threshold)

# 1) Naive MC: Z ~ N(0,1)
z_naive <- rnorm(n)
p_hat_naive <- mean(z_naive > threshold)

# 2) Importance sampling: sample from N(mu,1)
mu <- 4
z_is <- rnorm(n, mean = mu, sd = 1)

# weights: f/g = dnorm(z;0,1) / dnorm(z;mu,1)
w <- dnorm(z_is, mean = 0, sd = 1) / dnorm(z_is, mean = mu, sd = 1)

p_hat_is <- mean((z_is > threshold) * w)

c(p_true = p_true, naive = p_hat_naive, importance = p_hat_is)
```

A quick diagnostic: effective sample size (ESS)
===

Importance sampling can fail if weights are extremely variable.

A simple diagnostic is the **effective sample size**:
\[
\text{ESS} = \frac{\left(\sum_i w_i\right)^2}{\sum_i w_i^2}.
\]

- ESS close to \(n\): weights are stable
- ESS much smaller than \(n\): many weights are near zero and a few dominate

```{r, echo=TRUE}
ess <- (sum(w)^2) / sum(w^2)
c(n = n, ESS = ess, ESS_fraction = ess/n)
```

Intuition 
===

**Rejection Sampling**  
Think of a proposal distribution as your *street-corner strategy* for who you approach.

- You want a perfect sample of a very specific group (e.g., **left-handed violinists**).
- But instead of stopping *literally anyone*, you choose a *smart place to stand* (your **proposal**):
  - outside a music conservatory (better),
  - at a violin shop (even better),
  - vs. outside a random grocery store (terrible).

**Algorithm analogy:**
1. You pick someone according to your street-corner strategy (**proposal**).

2. You ask “Are you a left-handed violinist?”  

3. Even if they *seem* plausible, you still only accept them with a certain probability

4. Everyone you reject is thrown away.

**Takeaways:**
- If you choose a good proposal (stand near violinists), you accept often and it’s efficient.

- If you choose a bad proposal (random street), you reject almost everyone and waste time.

- The payoff: the accepted people are an **exact, unbiased sample** from your target group.

---

**Importance Sampling**  
Now you’re not trying to *literally isolate* left-handed violinists. You’re trying to estimate something about them (e.g., **average opinion**), efficiently.

- You again choose a smart place to stand (your **proposal**), e.g., “anyone who plays an instrument.”
- But now you **keep everyone’s answers**.
- Then you assign each person a **weight** that corrects for the fact that you sampled from an “instrument-player-heavy” crowd instead of the general population.

**Interpretation of weights:**
- A right-handed drummer might get a tiny weight (not very informative for left-handed violinists).

- A left-handed violinist gets a big weight (highly informative).

**Takeaways:**

- You don’t discard data, so it can be very efficient for *rare* groups/events.

- You’re typically estimating an expectation/probability about the target group, not producing a clean dataset of only target members.

- The risk: if your proposal still doesn’t cover the right kinds of people (or over/under-covers in the wrong way), weights can become extreme and the estimate gets unstable.

