---
title: 'Computational Methods for Hypothesis Testing and Sampling Distributions '
author: "Statistical Computing for Data Analysis"
output:
  slidy_presentation:
    font_adjustment: -1
  ioslides_presentation: default
---

Earlier Material
===

- how to simulate from distributions
- how to estimate expectations/probabilities
- how to sample efficiently

New question
===
Now we shift to using simulation for statistical inference.

How can we use repeated simulations (sometimes known as resampling) to carry out
**hypothesis testing** and understanding **sampling distributions**

- **Before**: simulate random *observations* or *data values* from a target distribution.
- **Now**: simulate *statistics*
- **Goal**: approximate the distribution of a (test) statistic. This distribution is known as a sampling distribution. 

What is a sampling distribution? 
===
Suppose we repeatedly drawn samples of size $n$ from the same population. Then for each sample, we compute a statistic, such as: 

- the sample mean $\bar{X}$
- the sample proportion $\hat{p}$
- the sample median
- the difference in two sample means


Note that **a statistic is random because the sample is random.**
Therefore it has a distribution. 
This distribution is called a **sampling distribution**. 

The **sampling distribution** is the distribution of that statistic over repeated samples.

Example: 
If we repeatedly sample $n = 25$ observations and compute $\bar{X}$ each time, the value of $\bar{X}$ will vary from sample to sample (i.e. the statistic $\bar{X}$ is random). 

The distribution of $\bar{X}$ is the sampling distribution of the sample mean. 

Population distribution vs. sampling distribution
===

**Population distribution**: Describes the distribution of individual data values or observations in the population. 

This is the distribution we were targeting when we used: 

- the inverse CDF method
- rejection sampling
- importance sampling or sampling importance resampling 
  
**Sampling distribution**: Describes the distribution of a statistic computed from repeated samples.


- Earlier: simulate **observations** from a population distribution
- Now: study the distribution of a **statistic** built from those observations


\[
\text{Population distribution}
\;\longrightarrow\;
\text{draw a sample } X_1,\dots,X_n
\;\longrightarrow\;
\text{compute a statistic } T(X_1,\dots,X_n)
\]

If we repeat this process many times, the resulting values of $T$ form the **sampling distribution** of the statistic.


Illustration
===
```{r}
set.seed(4279)

# Settings
n <- 25          # sample size
B <- 5000        # number of repeated samples

# Store the sample means
sample_means <- numeric(B)

# Repeated sampling
for (b in 1:B) {
  x <- rnorm(n, mean = 10, sd = 2)   # draw one sample from the population
  sample_means[b] <- mean(x)         # compute the statistic
}

# View the sampling distribution
hist(sample_means,
     main = "Sampling Distribution of the Sample Mean",
     xlab = "Sample mean",
     col = "lightgray",
     border = "white")

# Compare to the population mean
abline(v = 10, lwd = 2, lty = 2)
```


```{r}
set.seed(123)

# One sample from the population
x <- rnorm(25, mean = 10, sd = 2)

par(mfrow = c(1, 2))

hist(x,
     main = "One Sample of Raw Data",
     xlab = "x",
     col = "lightgray",
     border = "white")

hist(sample_means,
     main = "Sampling Distribution of x-bar",
     xlab = "Sample mean",
     col = "lightgray",
     border = "white")

abline(v = 10, lwd = 2, lty = 2)
```

Why do we care about sampling distributions?
===

Sampling distributions tell us how much a statistic varies from sample to sample. They help us answer questions like: 

- How much uncertainty is there in my estimate?
- How can I compute a standard error?
- How can I construct a confidence interval?
- How can I test a hypothesis?

Let's focus on this last bullet point. 

Now we understand sampling distributions, how do we use these sampling distributions to carry out statistical inference (i.e. hypothesis tests)?


From sampling distribution to inference
===
Suppose we want to compare two groups. A natural choice of statistic is

\[
T = \bar X_1 - \bar X_2
\]

which measures the difference in sample means. We could also standardize this $T$ by dividing by its standard error. 

We now ask: 

**Is this observed value large enough that
it suggests a real effect, or could it reasonably have occurred just by chance?** 

Let's say we calculate that $T$ is 0.5. Does this suggest a real effect? What if $T=2$ or $T= -1.5$. How large does $|T|$ need to be to suggest a real difference between $\bar X_1$ and $\bar X_2$? 

How do we answer this in a statistically principled way? 

From a statistic to a null distribution
===
Suppose we want to compare two groups, and we use

\[
T = \bar X_1 - \bar X_2
\]

to measure the difference between their sample means.

This quantity $T$ is called a **test statistic**.

Once we compute the observed value of $T$, the next question is:

**Is this difference large enough to suggest a real effect, or could it have happened just by chance?**

To answer that, we need a **reference distribution** for $T$.

In hypothesis testing, the reference distribution is the distribution of $T$ when the **null hypothesis is true**. 

This is called the **null distribution**.

### Think back

Up to this point in your statistics courses, how did you usually determine the null distribution?

Why the null distribution matters
===

Suppose our test statistic is

\[
T = \bar X_1 - \bar X_2
\]

and our observed value is

\[
T_{\text{obs}} = 4.2.
\]

By itself, the number $4.2$ does not tell us very much.

Its meaning depends on the reference distribution:

- if values around $4.2$ are common under the null, then this is not very surprising
- if values around $4.2$ are very rare under the null, then this is evidence against $H_0$.

Hypothesis testing workflow
===

The basic logic is:

1. Determine your null hypothesis and alternative hypothsis. 
2. Compute a test statistic (evidence from your data)
3. Determine its distribution under the null
4. Check whether the observed value is unusual (decision rule: p-value or rejection region)

### Guiding question

If the null hypothesis were true, how likely would it be to see a statistic this extreme?

This is what leads to:

- null distributions
- $p$-values
- rejection regions

Where do null distributions come from?
===

Sometimes we can derive the null distribution mathematically.

Examples you have seen before:

- normal distributions
- $t$ distributions
- chi-square distributions

But in many problems, the null distribution is difficult to obtain analytically.

That is where **simulation** and **resampling** become useful. 

Transition to resampling
===

If the null distribution is hard to derive, we can approximate it by repeated computation.

This leads naturally to methods such as:

- Monte Carlo simulation under $H_0$
- permutation tests

Later we will talk about general distributions needed for inference, which will include methods like jackknife and bootstrap 

Monte Carlo simulation under $H_0$
===

Recap: 

- A **sampling distribution** describes how a statistic varies over repeated samples.

- In hypothesis testing, we care about the sampling distribution of the test statistic **when the null hypothesis is true**.

- That sampling distribution is called the **null distribution**.

### Main idea

If we know what the data should look like under $H_0$, we can:

1. simulate many samples under $H_0$
2. compute the test statistic for each sample
3. use the resulting values to approximate the null distribution

Example: Monte Carlo null distribution
===

Suppose we want to test

\[
H_0: \mu = 10
\]

using the test statistic

\[
T = \bar X.
\]

If $H_0$ is true and the population is Normal with standard deviation $\sigma = 2$, then we can:

- repeatedly simulate samples from $N(10, 2^2)$
- compute $\bar X$ for each sample
- use the simulated values of $\bar X$ to approximate the null distribution of $T$

Then we compare our observed statistic to this simulated null distribution.

This is called **Monte Carlo simulation under $H_0$**.

```{r}
set.seed(123)

# Observed data
x_obs <- c(10.8, 9.7, 10.5, 11.1, 9.9, 10.4, 10.2, 10.7,
           9.8, 10.9, 10.1, 10.6, 11.0, 9.6, 10.3, 10.8,
           10.0, 10.5, 10.4, 10.7)

# Observed test statistic
T_obs <- mean(x_obs)

# Null hypothesis: mu = 10
mu0 <- 10
sigma <- 2
n <- length(x_obs)
B <- 5000

# Monte Carlo simulation under H0
T_null <- numeric(B)

for (b in 1:B) {
  x_sim <- rnorm(n, mean = mu0, sd = sigma)
  T_null[b] <- mean(x_sim)
}

# Plot simulated null distribution
hist(T_null,
     main = "Monte Carlo Approximation of the Null Distribution",
     xlab = "Simulated values of T = sample mean",
     col = "lightgray",
     border = "white")

abline(v = T_obs, lwd = 2, lty = 2)

# Two-sided Monte Carlo p-value
p_val <- mean(abs(T_null - mu0) >= abs(T_obs - mu0))
p_val
```
 When we do not know the null distribution
===

Sometimes we do **not** know the distribution of the data under $H_0$ well enough to simulate new datasets from a null model.

So instead of generating new data from a known null distribution, we use the **observed data** to construct a reference distribution.

### Main idea

Under the null hypothesis, the observed data often satisfy some kind of **exchangeability** or **relabeling symmetry**.

This means that, if $H_0$ is true, certain rearrangements of the data should not change the distribution of the test statistic under $H_0$.

So we can:

1. keep the observed data values fixed
2. rearrange or reassign part of the data in a way justified by $H_0$
3. recompute the test statistic each time
4. use the resulting values to approximate the null distribution

This gives a **permutation distribution**, which is a data-based approximation to the null distribution.
### Examples

- In a two-sample problem, we may permute (shuffle) **group labels**
- In a correlation problem, we may permute one variable relative to the other
- In a regression problem, we may permute the response relative to the predictor

Permutation as a data-based null distribution
===

Suppose we want to test whether two groups differ, using

\[
T = \bar X_1 - \bar X_2.
\]

If we do **not** know the data distribution under $H_0$, we need another way to approximate the null distribution of $T$.

In this two-sample test setting, we are testing: 
$H_0: \mu_1 = \mu_2$ versus $H_1: \mu_1 \neq \mu_2$. So under the null hypothesis of **no group difference**, the group labels are interchangeable.

So we:

- keep the observed data values fixed
- permute (shuffle) the labels many times
- recompute $T$ after each shuffle

The resulting distribution of these shuffled statistics is the **permutation distribution**.

We use this as a data-based null distribution for hypothesis testing.